Computational biology in the 21st century: Algorithms that scale

### 2015-07-28 main.tex
treepoints figure, fix figure margins

\usepackage{caption}
\usepackage{subfig}

@@

% Page heads
\markboth{B. Berger et al.}{Title}

% Title portion
\title{Title}
\author{BONNIE BERGER
\affil{Massachusetts Institute of Technology}
NOAH M. DANIELS
\affil{Massachusetts Institute of Technology}
Y. WILLIAM YU
\affil{Massachusetts Institute of Technology}}

\begin{abstract}
abstract goes here
\end{abstract}

@@

\begin{bottomstuff}
% This work is supported by the National Science Foundation, under
% grant CNS-0435060, grant CCR-0325197 and grant EN-CS-0329609.
%
% Author's addresses: G. Zhou, Computer Science Department,
% College of William and Mary; Y. Wu  {and} J. A. Stankovic,
% Computer Science Department, University of Virginia; T. Yan,
% Eaton Innovation Center; T. He, Computer Science Department,
% University of Minnesota; C. Huang, Google; T. F. Abdelzaher,
% (Current address) NASA Ames Research Center, Moffett Field, California 94035.
\end{bottomstuff}

@@

\begin{figure}[htb!]
\centering
\subfloat[Growth of genomic sequence data (orange) as compared with the combined power of the top-500 supercomputer list (blue); y-axis is logarithmic]{
\includegraphics[width=3.0in]{assets/moore.png}
\label{fig:moore}
}\\
\subfloat[Growth of genomic sequence data (orange) as compared with hard drive capacity (blue); y-axis is logarithmic]{
\includegraphics[width=3.0in]{assets/kryder.png}
\label{fig:kryder}
}
\caption[Sequencing vs. Storage]{Moore's (a) and Kryder's (b) laws contrasted with genomic sequence data}
\label{fig:growth}
\end{figure}

@@

Given DNA or RNA reads from NGS technologies, the first task is to 
\emph{assemble} those fragments of sequence (which have associated quality 
scores) into contiguous sequences.
The task of \emph{de novo} assembly is beyond the scope of this Review, but is
possible because reads overlap and because the sequence is covered by many such
overlapping reads~\cite{zhang2011practical}; the \emph{de Bruijn graph} is a 
commonly-used data structure.
Often, however, a \emph{reference genome} (or in the case of RNA, 
\emph{transcriptome}) is available for the organism being sequenced; the
establishment of a human reference genome was indeed the purpose of the Human
Genome Project.
When a reference sequence is available, NGS reads can be \emph{mapped} onto
this reference (Figure~\ref{fig:ngs-pipeline}).
Mapping allows the differences between the newly-sequenced genome and the reference
to be analyzed; these differences may include single-nucleotide polymorphisms
(SNPs, which are the genetic analogue to bit-flips) (Figure~\ref{fig:snps}), 
insertions or deletions, or larger-scale changes in the genome.
While reference-based read mapping is a fundamentally simpler problem than
\emph{de novo} assembly, it is still computationally complex, as gigabytes or
terabytes of reads must each be mapped onto the reference genome, which can 
range from millions (for bacteria) to billions (for mammals).
As read mapping is typically the most costly step in NGS analysis pipelines (e.g., GATK), any improvement to existing mappers will immediately accelerate sequence analysis studies on large read datasets.

\begin{figure}[htb!]
\centering
\subfloat[``Shotgun'' sequencing breaks DNA molecules into many short fragments, or reads, and relies on high coverage to produce a statistically likely representation of a whole genome]{
\includegraphics[width=3.0in]{assets/shotgun.png}
\label{fig:shotgun}
}
\subfloat[Single-nucleotide polymorphisms, or SNPs, are the simplest type of genomic variant, and form the bulk of ``variant-calling'' analysis]{
\includegraphics[width=3.0in]{assets/snps.png}
\label{fig:snps}
}\\
\subfloat[The NGS downstream analysis pipeline. Shotgun reads are mapped to a reference genome with tools such as BWA or Bowtie. The resulting genomic sequence is analyzed for variants with tools such as GATK or Samtools. This allows relationships between genes and diseases to be uncovered.]{
\includegraphics[width=3.0in]{assets/ngs-pipeline.png}
\label{fig:ngs-pipeline}
}
\caption[The NGS Pipeline]{The next-generation sequencing (NGS) pipeline.}
\label{fig:ngs}
\end{figure}

@@

Given a sequenced genome, the next natural questions are what genes (genomic
regions that code for proteins) are present, what structure each resulting
protein takes, and what biological function it performs.
Identifying likely genes is a well-studied problem~\cite{burge1998finding} 
beyond the scope of this Review.
However, determining evolutionary relationships, structure, and function are
at the heart of current research in computational biology.
Since some organisms (known as \emph{model organisms}) are better studied than
others, and evolution is known to conserve sequence, structure, and function,
a powerful approach to determine these attributes is to search for similar
sequences about which more is known. 
This so-called \emph{homology search} entails searching for \emph{approximate}
matches in databases of known gene or protein sequences.
The homology search problem was previously believed to be solved;
Basic Local Alignment Search Tool (BLAST)~\cite{altschul1990basic} had been
the standard tool for performing homology (similarity) search on databases of nucleotide and
protein sequences.
BLAST takes a ``seed-and-extend'' approach; it looks for small, $k$-mer matches
that might lead to longer matches, and greedily extends them, ultimately 
producing a \emph{sequence alignment} between a query and each potential 
database hit.
However, BLAST's running time scales linearly with the size of the database being
searched, which is problematic as sequence databases continue to grow at a
faster rate than Moore's law.

@@

Over the course of evolutionary time, structure is known to be more highly 
conserved than sequence~\cite{illergaard2009structure}, which is to say that it 
does not change as rapidly.
When the structure of a protein is known but its biological function or 
evolutionary relationships are not, researchers may search for structurally
similar proteins that are better studied~\cite{gibrat1996surprising}.
Classical tools for this involve performing pairwise structural alignments to
look for geometric similarity; DALI~\cite{holm1995dali} is still widely used,
along with other aligners such as FATCAT~\cite{ye2004fatcat} and 
Matt~\cite{menke2008matt}.
Due to the complexity of protein structures, these programs generally take 
significant amounts of time,
especially to align multiple structures.
For example, the DALI webserver can take as much as an hour to return results
for a single query.
FragBag~\cite{budowski2010fragbag} accelerates protein structure search by
approximating structural alignments
by instead comparing the ``bag-of-words'' from each structure.
Analogous to a term-frequency vector in information retrieval, this bag-of-words
indicates the abundance of particular, short structural motifs within a protein.
%Does Fragbag have challenges we should mention?
% NMD: not really. That we speed it up is evidence of the orthogonality of
% our approach

The study of drugs and their targets based on chemical structure and function
is known as chemogenomics~\cite{bredel2004chemogenomics}.
In the fields of drug discovery and drug repurposing, the prediction 
of biologically active compounds is an important task. 
Computational high-throughput screening eliminates many compounds from 
laborious wet-lab consideration, but even computational screening can be 
time-consuming.
Notably, large chemical compound databases, such as PubCHEM,
cannnot be searched on a laptop computer with standard tools (e.g., SMSD).

@@

Techniques for reference-based read mapping typically rely on algorithmic 
approaches such as the Burrows-Wheeler transform (BWT)~\cite{burrows1994block}, 
which provides efficient string 
compression through a reversible transformation, while the FM-index data 
structure~\cite{ferragina2000opportunistic} is a compressed substring index, 
based on the BWT, which provides efficient storage as well as fast search.
BWA (Burrows-Wheeler Aligner)~\cite{li2009fast} uses the BWT, while the 
Bowtie2~\cite{langmead2012fast} mapper further relies on the FM-index for 
efficient mapping of NGS reads.
The Genome Multitool (GEM) mapper~\cite{marco2012gem} also uses an FM-index 
coupled with dynamic programming in a compressed representation of the 
reference genome, in order to prune the search space 
when mapping reads to a reference genome.
Masai~\cite{siragusa2013fast} uses an ``approximate seed'' approach to indexing
the space of possible matches, likewise pruning the search space; however, the
bulk of its runtime is spent on the extend phase.
Even with these approaches, read mapping remains a significant bottleneck in
genomic research~\cite{hach2010mrsfast}.

Recent advances in metagenomic search tools have relied on two advances over
BLASTX: indexing and alphabet reduction.
RapSearch2~\cite{zhao2012rapsearch2} relies on alphabet reduction and a 
collision-free hash table.
The alphabet reduction, as it is reversible, can be thought of as a form of
lossless compression; a 20-letter amino acid alphabet is mapped onto a smaller
alphabet, with offsets stored to recover the original sequence in the full
alphabet.
The hash-table provides an efficient index of the database to be searched.
DIAMOND~\cite{buchfink2014fast} also relies on alphabet reduction, but uses
``shaped seeds'' instead of simple $k$-mer seeds to index the database.
DIAMOND demonstrates search performance three to four orders of magnitude faster
than BLASTX, but it is still linear in the size of the database being searched.

@@

An area of great interest in integrating network data is in global multiple network 
alignment, which is based on network topology and conserved biological function (IsoRank~\cite{singh2008global} and IsoRankN~\cite{liao2009isorankn}).
IsoRank's approach is similar to the PageRank algorithm~\cite{page1999pagerank} 
but performs a random walk on the product graph of two networks;
IsoRankN produces a \emph{Personalized PageRank} alignment, similar to the
PageRank-Nibble algorithm~\cite{andersen2006local}.
Diffusion state distance (DSD)~\cite{cao2013going} is a diffusion-based approach
which defines a distance metric between two nodes based on random walks.
DSD penalizes high-degree nodes, and outperforms other graph-theoretic distance
functions when applied to standard approaches such as neighbor voting for function annotation.
Diffusion component analysis (DCA)~\cite{cho2015diffusion} amounts to a network-based 
analogue of principle component analysis; DCA reduces the dimensionality of a
network in such a way that function annotation can be performed in the reduced 
space, significantly decreasing running time while still providing competitive
prediction accuracy.
Using a compressed representation of metabolic networks, 
Mongoose~\cite{chindelevitch2014exact} is able to 
identify and unblock many blocked reactions in current models without loss of 
accuracy.

\section{Compressive omics}

Another approach is to make use of a completely different paradigm for the 
structure of biological data.
Much biological data is highly redundant; if a computation is performed on one
human genome, and a researcher wishes to perform the same computation on another
human genome, most of the work has already been done.

Seeking to take advantage of the redundancy inherent in genomic sequence data, 
Loh, Baym and Berger \cite{loh2012compressive}
introduced \emph{compressive genomics}, an approach that relies on compressing
data in such a way that the desired computation (such as BLAST search) can be
performed in the compressed representation.
This approach relies on a two-stage search, referred to as \emph{coarse} and \emph{fine} search.
Coarse search is performed only on the coarse, or representative, subsequences
that represent unique data.
Any representative sequence that is within some threshold of the query is then expanded 
into all the similar sequences it represents; the fine search is then over this
(typically small) subset of the original database.
This approach provides run-time improvements to BLAST search on 
nucleotide~\cite{loh2012compressive} and protein~\cite{daniels2013compressive}
data; moreover, these run-time improvements increase as databases grow.

The CORA read mapper~\cite{yorukoglu2015compressive} applies an $l$-mer based 
read-compression approach with a compressive indexing of the reference genome
(referred to as a homology table).
CORA, like caBLAST and caBLASTP, accelerates existing tools (in this case, read
mappers including BWA, and Bowtie2) by allowing them to operate in a compressed
space, and relying on a coarse and a fine phase.
Short seed-clustering schemes, such as those used in Masai~\cite{siragusa2013fast} and 
MrsFAST~\cite{hach2010mrsfast} conceptually differ from CORA in that those 
schemes aim to accelerate only the seed-to-reference matching step.
Thus, there is a subsequent seed-extension step, which is substantially more 
costly and still needs to be performed for each read and mapping individually, 
even when seeds are clustered. 
Through its $l$-mer based read compression model, CORA is able to accelerate 
and 
achieve asymptotically sublinear scaling for both the seed-matching and 
seed-extension steps within coarse-mapping, which comprises the major bulk of 
the read-mapping computation.
Traditionally, $k$-mers refer to short substrings of fixed length (often, but 
not necessarily, a power of two) used as ``seeds'' for longer sequence matches.
CORA uses much longer $k$-mers, and links each one to its neighbors within
a small Hamming or Levenshtein distance.
The term $l$-mer distinguishes these substrings from typically-short $k$-mers.

In the area of metagenomic search,
the recently-released MICA~\cite{yu2015entropy} demonstrates that the 
compressive-acceleration approach of 
caBLAST~\cite{loh2012compressive} and caBLASTP~\cite{daniels2013compressive} is 
largely orthogonal to alphabet-reduction and indexing approaches.
MICA applies the compressive-acceleration framework to 
DIAMOND~\cite{buchfink2014fast}, using DIAMOND
for its ``coarse search'' phase and a user's choice of DIAMOND or BLASTX for its
``fine search'' phase; it demonstrates comparable run-time gains over DIAMOND as
caBLASTP does over BLASTP.

The compressive acceleration approach of \cite{loh2012compressive} has been
generalized and adapted to non-sequence spaces as well.
Chemogenomics, the computational study of small-molecule compounds and their
affect on phenotype, relies on comparing chemical graph structures to identify
similar molecules and binding sites.
However, there are once again an increasing number of such chemical compounds
to search; the NCBI's PubChem database has grown from 31 million compounds in
January 2011 to 68 million in July, 2015.
Furthermore, comparing chemical graph structures typically involves computing
the maximal common subgraph (MCS), an NP-hard problem.
One approach to this type of search is the Small Molecule Subgraph Detector 
(SMSD)~\cite{rahman2009small}, which applies one of several MCS algorithms based
on the size and complexity of the graphs in question.
Applying a compressive acceleration approach, Ammolite~\cite{yu2015entropy}
accelerates SMSD search by an average of 150x on the PubChem database.

Again applying a compressive acceleration approach, 
esFragBag~\cite{yu2015entropy} clusters proteins based on the cosine distance or
Euclidean distance of their bag-of-words vectors, further accelerating FragBag's
running time by an average of 10x.

\section{Structure of biological data}

Cluster-based search is well-studied~\cite{jardine1971use}, but conventional
wisdom holds that it provides a constant factor speed-up over exhaustive search.
However, \cite{daniels2013compressive} observed that caBLASTP demonstrated
increased gains over BLASTP on newer, larger protein databases, suggesting that
compressive acceleration takes advantage of ever-increasing redundancy in the
data.
Beyond redundancy, however, another attribute of large biological data sets
stands out.
Any high-dimensional data set, such as amino acid sequences of a bounded 
length, can be described as inhabiting a vector space.
For example, given an alphabet of 20 amino acids and a length bounded at 
500\footnote{While the largest-known protein, titin, is over 33,000 amino acids 
in length, proteins are typically described as comprising \emph{domains} that 
evolve and fold independently; domains rarely exceed 500 amino acids.}, one 
could
enumerate $20^{500} \approx 3 \times 10^{650}$ distinct protein sequences, as 
of July 
2015, the NCBI's exhaustive ``NR'' database lists $6.9 \times 10^{7}$ sequences.
This is not simply due to undersampling; thanks to evolution, only those protein
sequences which exhibit useful biological function should survive, and most
random sequences of amino acids would not be expected to form stable structures.
Thus, not only do biological data exhibit redundancy, they also tend not to 
inhabit the entire feasible space.
It seems that physical laws have constrained the data to a particular subspace
of the enumerable Cartesian space.

One key insight as to why the ``compressive omics'' approach allows computation
that scales sublinearly in the size of the data is that such data sets exhibit
low \emph{metric entropy}~\cite{yu2015entropy}.
That is, for a given cluster radius $r_c$ and a database $D$, the number $k$ 
of clusters needed to cover $D$ is bounded by $N_{r_c} (D)$, the metric 
entropy, which is relatively small compared to
$|D|$, the number of entries in the database.
In contrast, if the points were uniformly distributed about the Cartesian space,
$N_{r_c} (D)$ would be larger.

A second key insight is that the biological data sets considered 
by~\cite{yu2015entropy} exhibit low \emph{fractal dimension}.
That is, within some range of radii $r_1$ and $r_2$ about an arbitrary point
in the database $D$, the fractal dimension $d$ is
$d=\frac{\log (n_2 / n_1)}{ \log (r_2 / r_1)}$, where $n_1$ and $n_2$ are the
number of points within $r_1$ and $r_2$ respectively.

@@

Biological data sets exhibit redundancy, and are constrained to subspaces by
physical laws.
This combination results in low fractal dimension and low metric entropy 
relative to the size of the data set, which suggests that ``compressive omics''
will provide the ability for computation to scale with massively growing data.

@@

@@

>>>>

\usepackage{caption}
\usepackage{subfig}
\captionsetup[subfigure]{margin=5pt}

@@

% Page heads
\markboth{B. Berger et al.}{Computational biology in the 21st century: Algorithms that scale}

% Title portion
\title{Computational biology in the 21st century: Algorithms that scale}
\author{BONNIE BERGER
\affil{Massachusetts Institute of Technology}
NOAH M. DANIELS
\affil{Massachusetts Institute of Technology}
Y. WILLIAM YU
\affil{Massachusetts Institute of Technology}}

\begin{abstract}
Throughout all areas of data science, scientists are being confronted by a
veritable explosion of available data. In many fields, this increase is
exponential in nature, even outpacing Moore's and Kryder's laws on the
respective doublings of transistors on a chip and long-term data storage
density. As such, the challenges posed by the massive influx of data cannot
be solved by waiting for faster and larger capacity computers, but require
instead the development of data structures and representations that exploit
and simplify complexity in the dataset. In this Review, we
survey the biological data landscape, with a focus on how taking advantage
of the structure of massive biological data enables the design of algorithms
that scale sublinearly in large-scale genomics, personal genomics and
chemogenomics.
\end{abstract}

@@

\begin{bottomstuff}
This work is supported by the National Institutes of Health, under
grant GM108348.
% This work is supported by the National Science Foundation, under
% grant CNS-0435060, grant CCR-0325197 and grant EN-CS-0329609.
%
% Author's addresses: G. Zhou, Computer Science Department,
% College of William and Mary; Y. Wu  {and} J. A. Stankovic,
% Computer Science Department, University of Virginia; T. Yan,
% Eaton Innovation Center; T. He, Computer Science Department,
% University of Minnesota; C. Huang, Google; T. F. Abdelzaher,
% (Current address) NASA Ames Research Center, Moffett Field, California 94035.
\end{bottomstuff}

@@

\begin{figure}[htb!]
\centering
\subfloat[Growth of genomic sequence data (orange) as compared with the combined power of the top-500 supercomputer list (blue); y-axis is logarithmic]{
\includegraphics[width=2.8in]{assets/moore.png}
\label{fig:moore}
}
\subfloat[Growth of genomic sequence data (orange) as compared with hard drive capacity (blue); y-axis is logarithmic]{
\includegraphics[width=2.8in]{assets/kryder.png}
\label{fig:kryder}
}
\caption[Sequencing vs. Storage]{Moore's (a) and Kryder's (b) laws contrasted with genomic sequence data}
\label{fig:growth}
\end{figure}

@@

Given DNA or RNA reads from NGS technologies, the first task is to 
\emph{assemble} those fragments of sequence (which have associated quality 
scores) into contiguous sequences.
The task of \emph{de novo} assembly is beyond the scope of this Review, but is
possible because reads overlap and because the sequence is covered by many such
overlapping reads~\cite{zhang2011practical}; the \emph{de Bruijn graph} is a 
commonly-used data structure.
Often, however, a \emph{reference genome} (or in the case of RNA, 
\emph{transcriptome}) is available for the organism being sequenced; the
establishment of a human reference genome was indeed the purpose of the Human
Genome Project.

When a reference sequence is available, NGS reads can be \emph{mapped} onto
this reference (Figure~\ref{fig:ngs-pipeline}).
Mapping allows the differences between the newly-sequenced genome and the 
reference to be analyzed; these differences, or variants, may include single-nucleotide polymorphisms
(SNPs, which are the genetic analogue to bit-flips) (Figure~\ref{fig:snps}), 
insertions or deletions, or larger-scale changes in the genome.
Determining the differences between an individual genome and a reference is
known as \emph{variant calling}.
While reference-based read mapping is a fundamentally simpler problem than
\emph{de novo} assembly, it is still computationally complex, as gigabytes or
terabytes of reads must each be mapped onto the reference genome, which can 
range from millions (for bacteria) to billions (for mammals).
As an example, the ICGC-TCGA Pan Cancer Analysis of Whole Genomes 
(PCAWG)~\cite{weinstein2013cancer} brings together more than 500 top cancer researchers from about 80 institutions in a coordinated manner with the goal of mapping the entire mutational landscape of 37 common cancer types. 
Currently each sample requires seven hours to download. 
Importantly, researchers do not trust the provided mapping, and thus they redo mappings. 
The time spent on mapping is about 50\% of the overall time spent on the sequence analysis pipeline. 
As read mapping is typically the most costly step in NGS analysis pipelines (e.g., GATK), any improvement to existing mappers will immediately accelerate sequence analysis studies on large read datasets.

\begin{figure}[htb!]
\centering
\subfloat[``Shotgun'' sequencing breaks DNA molecules into many short fragments, which are read from one or both ends in the form of {\em reads,\/} and relies on high coverage to produce a statistically likely representation of a whole genome]{
\includegraphics[width=2.6in]{assets/shotgun.png}
\label{fig:shotgun}
}
\subfloat[Single-nucleotide polymorphisms, or SNPs, are the simplest type of genomic variant, and form the bulk of ``variant-calling'' analysis]{
\includegraphics[width=2.6in]{assets/snps.png}
\label{fig:snps}
}\\
\subfloat[The NGS downstream analysis pipeline. Shotgun reads are mapped to a reference genome with tools such as BWA or Bowtie. The resulting genomic sequence is analyzed for variants with tools such as GATK or Samtools. This allows relationships between genes and diseases to be uncovered.]{
\includegraphics[width=4.0in]{assets/ngs-pipeline.png}
\label{fig:ngs-pipeline}
}
\caption[The NGS Pipeline]{The next-generation sequencing (NGS) pipeline.}
\label{fig:ngs}
\end{figure}

@@

Given a sequenced genome, the next natural questions are what genes (genomic
regions that code for proteins) are present, what structure each resulting
protein takes, and what biological function it performs.
Identifying likely genes is a well-studied problem~\cite{burge1998finding} 
beyond the scope of this Review.
However, determining evolutionary relationships, structure, and function are
at the heart of current research in computational biology.
Since some organisms (known as \emph{model organisms}) are better studied than
others, and evolution is known to conserve sequence, structure, and function,
a powerful approach to determine these attributes is to search for similar
sequences about which more is known. 
This so-called \emph{homology search} entails searching for \emph{approximate}
matches in databases of known gene or protein sequences.
The homology search problem was previously believed to be solved;
Basic Local Alignment Search Tool (BLAST)~\cite{altschul1990basic} had been
the standard tool for performing homology (similarity) search on databases of nucleotide and
protein sequences.
BLAST takes a ``seed-and-extend'' approach; it looks for small, $k$-mer matches
that might lead to longer matches, and greedily extends them, ultimately 
producing a \emph{sequence alignment} between a query and each potential 
database hit.
However, BLAST's running-time scales linearly with the size of the database being
searched, which is problematic as sequence databases continue to grow at a
faster rate than Moore's law.

@@

Over the course of evolutionary time, structure is known to be more highly 
conserved than sequence~\cite{illergaard2009structure}, which is to say that it 
does not change as rapidly.
When the structure of a protein is known but its biological function or 
evolutionary relationships are not, researchers may search for structurally
similar proteins that are better studied~\cite{gibrat1996surprising}.
Classical tools for this involve performing pairwise structural alignments to
look for geometric similarity; DALI~\cite{holm1995dali} is still widely used,
along with other aligners such as FATCAT~\cite{ye2004fatcat} and 
Matt~\cite{menke2008matt}.
Due to the complexity of protein structures, these programs generally take 
significant amounts of time,
especially to align multiple structures.
For example, the DALI webserver can take as much as an hour to return results
for a single query.
FragBag~\cite{budowski2010fragbag} accelerates protein structure search by
approximating structural alignments
by instead comparing the ``bag-of-words'' from each structure.
Analogous to a term-frequency vector in information retrieval, this bag-of-words
indicates the abundance of particular, short structural motifs within a protein.

The computational study of drugs and their targets based on chemical structure and function
is known as chemogenomics~\cite{bredel2004chemogenomics}.
In the fields of drug discovery and drug repurposing, the prediction 
of biologically active compounds is an important task. 
Computational high-throughput screening eliminates many compounds from 
laborious wet-lab consideration, but even computational screening can be 
time-consuming.
Chemogenomics typically relies on comparing chemical graph structures to identify
similar molecules and binding sites.
Furthermore, comparing chemical graph structures typically involves computing
the maximal common subgraph (MCS), an NP-hard problem.
However, there are once again an increasing number of such chemical compounds
to search; the NCBI's PubChem database has grown from 31 million compounds in
January 2011 to 68 million in July, 2015.

@@

Techniques for reference-based read mapping typically rely on algorithmic 
approaches such as the Burrows-Wheeler transform (BWT)~\cite{burrows1994block}, 
which provides efficient string 
compression through a reversible transformation, while the FM-index data 
structure~\cite{ferragina2000opportunistic} is a compressed substring index, 
based on the BWT, which provides efficient storage as well as fast search.
BWA (Burrows-Wheeler Aligner)~\cite{li2009fast} uses the BWT, while the 
Bowtie2~\cite{langmead2012fast} mapper further relies on the FM-index for 
efficient mapping of NGS reads.
The Genome Multitool (GEM) mapper~\cite{marco2012gem} also uses an FM-index 
coupled with dynamic programming in a compressed representation of the 
reference genome, in order to prune the search space 
when mapping reads to a reference genome.
Masai~\cite{siragusa2013fast} and mrsFAST \cite{hach2014mrsfast} use an ``approximate seed'' approach to indexing
the space of possible matches, likewise pruning the search space; however, the
bulk of its runtime is spent on the extend phase.
State-of-the-art mapper mrsFAST-Ultra achieves improvements in efficiency based on machine architecture rather than leveraging redundancy in the data itself with near-perfect sensitivity, but
only for the case where there are no insertions and deletions (indels) \cite{hach2014mrsfast}.
Even with these approaches, read mapping remains a significant bottleneck in
genomic research~\cite{hach2010mrsfast}.

Recent advances in metagenomic search tools have relied on two improvements over
BLASTX: indexing and alphabet reduction.
RapSearch2~\cite{zhao2012rapsearch2} relies on alphabet reduction and a 
collision-free hash table.
The alphabet reduction, as it is reversible, can be thought of as a form of
lossless compression; a 20-letter amino acid alphabet is mapped onto a smaller
alphabet, with offsets stored to recover the original sequence in the full
alphabet.
The hash-table provides an efficient index of the database to be searched.
DIAMOND~\cite{buchfink2014fast} also relies on alphabet reduction, but uses
``shaped seeds'' instead of simple $k$-mer seeds to index the database.
DIAMOND demonstrates search performance three to four orders of magnitude faster
than BLASTX, but it is still linear in the size of the database being searched.

@@

The most widely-used chemogenomics search is the Small Molecule Subgraph Detector 
(SMSD)~\cite{rahman2009small}, which applies one of several MCS algorithms based
on the size and complexity of the graphs in question.
Notably, large chemical compound databases, such as PubCHEM,
cannnot be searched on a laptop computer with current tools such as SMSD.

An area of great interest in integrating biological network data is in global multiple network 
alignment, which is based on network topology and conserved biological function (IsoRank~\cite{singh2008global} and IsoRankN~\cite{liao2009isorankn}).
IsoRank's approach is similar to the PageRank algorithm~\cite{page1999pagerank} 
but performs a random walk on the product graph of two networks;
IsoRankN produces a \emph{Personalized PageRank} alignment, similar to the
PageRank-Nibble algorithm~\cite{andersen2006local}.
Diffusion state distance (DSD)~\cite{cao2013going} is a diffusion-based approach
which defines a distance metric between two nodes based on random walks.
DSD penalizes high-degree nodes, and outperforms other graph-theoretic distance
functions when applied to standard approaches such as neighbor voting for function annotation.
Diffusion component analysis (DCA)~\cite{cho2015diffusion} amounts to a network-based 
analogue of principle component analysis; DCA reduces the dimensionality of a
network in such a way that function annotation can be performed in the reduced 
space, significantly decreasing running time while still providing competitive
prediction accuracy.
Using a compressed representation of metabolic networks, 
Mongoose~\cite{chindelevitch2014exact} is able to 
identify and unblock many blocked reactions in current models without loss of 
accuracy.

\section{Structure of biological data}\label{structure}

Fortunately, biological data has unique structure \cite{yu2015entropy}, which we 
will take advantage of to perform search that scales sublinearly in the size of the database~(Section~\ref{omics}).
The first critical observation is that
much biological data is highly redundant; if a computation is performed on one
human genome, and a researcher wishes to perform the same computation on another
human genome, most of the work has already been done \cite{loh2012compressive}.
When dealing with redundant data, clustering comes to mind.
While cluster-based search is well-studied~\cite{jardine1971use}, conventional
wisdom holds that it provides a constant factor speed-up over exhaustive search.

Beyond redundancy, however, another attribute of large biological data sets
stands out.
Any high-dimensional data set, such as amino acid sequences of a bounded 
length, can be described as inhabiting a vector space.
For example, given an alphabet of 20 amino acids and a length bounded at 
500\footnote{While the largest-known protein, titin, is over 33,000 amino acids 
in length, proteins are typically described as comprising \emph{domains} that 
evolve and fold independently; domains rarely exceed 500 amino acids.}, one 
could
enumerate $20^{500} \approx 3 \times 10^{650}$ distinct protein sequences, as 
of July 
2015, the NCBI's exhaustive ``NR'' database lists $6.9 \times 10^{7}$ sequences.
This surprisingly lower number is not simply due to undersampling; thanks to 
evolution, only those protein
sequences which exhibit useful biological function should survive, and most
random sequences of amino acids would not be expected to form stable structures.
Thus, not only do biological data exhibit redundancy, they also tend not to 
inhabit anywhere near the entire feasible space (Figure~\ref{fig:clusters}).
It seems that physical laws have constrained the data to a particular subspace
of the enumerable Cartesian space.

\begin{figure}[htb!]
\centering
\subfloat[Cartoon depiction of points in an arbitrary high-dimensional space, 
as might arise from genomes generated by mutation and selection during the 
course of evolution. 
Although high-dimensional locally, at the global scale of covering spheres, the 
data cloud looks nearly 1-dimensional, which enables entropy-scaling of 
similarity search. Clusters cover the data points but do not cover unoccupied 
regions of space.]{
\includegraphics[width=2.8in]{assets/treepoints.pdf}
\label{fig:clusters}
}
\subfloat[A query (green triangle) and two concentric search radii (red 
circles) around it. Thanks to low fractal dimension, the large circle does
not contain vastly more points than the small circle.]{
\includegraphics[width=2.8in]{assets/treepoints-fractal.png}
\label{fig:fractal}
}
\caption[Data spaces.]{A cartoon illustration of a biological data space.}
\label{fig:dataspace}
\end{figure}

One key insight related to redundancy
%as to why the compressive omics approach allows computation 
%that scales sublinearly in the size of the data 
is that such data sets exhibit
low \emph{metric entropy}~\cite{yu2015entropy}.
That is, for a given cluster radius $r_c$ and a database $D$, the number $k$ 
of clusters needed to cover $D$ is bounded by $N_{r_c} (D)$, the metric 
entropy, which is relatively small compared to
$|D|$, the number of entries in the database (Figure~\ref{fig:clusters}).
In contrast, if the points were uniformly distributed about the Cartesian space,
$N_{r_c} (D)$ would be larger.

A second key insight is that the biological data sets have low \emph{fractal dimension}~\cite{yu2015entropy}.
That is, within some range of radii $r_1$ and $r_2$ about an arbitrary point
in the database $D$, the fractal dimension $d$ is
$d=\frac{\log (n_2 / n_1)}{ \log (r_2 / r_1)}$, where $n_1$ and $n_2$ are the
number of points within $r_1$ and $r_2$ respectively (Figure~\ref{fig:fractal}).

@@

Biological data sets exhibit redundancy, and are constrained to subspaces by
physical laws.
This combination results in low fractal dimension and low metric entropy 
relative to the size of the data set, which suggests that ``compressive omics''
will provide the ability for computation to scale sublinearly with massively growing data.

@@

\section{The age of compressive omics}\label{omics}

We are entering the age of compressive omics, 
which makes use of this completely different paradigm for the structure of biological data.
Seeking to take advantage of the redundancy inherent in genomic sequence data, 
Loh, Baym and Berger \cite{loh2012compressive}
introduced \emph{compressive genomics}, an approach that relies on compressing
data in such a way that the desired computation (such as BLAST search) can be
performed in the compressed representation.
Daniels et al. \cite{daniels2013compressive} further demonstrated that
compressive acceleration takes advantage of increasing redundancy in the
data over time.

Compressive genomics is based on the concept of \emph{compressive acceleration},
which relies on a two-stage search, referred to as \emph{coarse} and \emph{fine} search.
Coarse search is performed only on the coarse, or representative, subsequences
that represent unique data.
Any representative sequence that is within some threshold of the query is then expanded 
into all the similar sequences it represents; the fine search is then over this
(typically small) subset of the original database.
This approach provides orders-of-magnitude run-time improvements to BLAST search on 
nucleotide~\cite{loh2012compressive} and protein~\cite{daniels2013compressive}
data; moreover, these run-time improvements increase as databases grow.

The CORA read mapper~\cite{yorukoglu2015compressive} applies a mid-size $l$-mer based 
read-compression approach with a compressive indexing of the reference genome
(referred to as a homology table).
CORA, like caBLAST and caBLASTP, accelerates existing tools (in this case, read
mappers including BWA or Bowtie2) by allowing them to operate in a compressed
space, and relying on a coarse and a fine phase.
In contrast, short seed-clustering schemes, such as those used in Masai~\cite{siragusa2013fast} and 
MrsFAST~\cite{hach2010mrsfast}, conceptually differ from CORA in that those 
schemes aim to accelerate only the seed-to-reference matching step.
Thus, there is a subsequent seed-extension step, which is substantially more 
costly and still needs to be performed for each read and mapping individually, 
even when seeds are clustered. 
Through its $l$-mer based read compression model, CORA is able to accelerate 
and 
achieve asymptotically sublinear scaling for both the seed-matching and 
seed-extension steps within coarse-mapping, which comprises the major bulk of 
the read-mapping computation.
Traditionally, $k$-mers refer to short substrings of fixed length (often, but 
not necessarily, a power of two) used as ``seeds'' for longer sequence matches.
CORA uses much longer $k$-mers (e.g., 33-64 nucleotides long), and links each one to its neighbors within
a small Hamming or Levenshtein distance.
The term $l$-mer distinguishes these substrings from typically-short $k$-mers.

In the area of metagenomic search,
the recently-released MICA~\cite{yu2015entropy} demonstrates that the 
compressive-acceleration approach of 
caBLAST~\cite{loh2012compressive} and caBLASTP~\cite{daniels2013compressive} is 
largely orthogonal to alphabet-reduction and indexing approaches.
MICA applies the compressive-acceleration framework to the
state-of-the-art DIAMOND~\cite{buchfink2014fast}, using it
for its ``coarse search'' phase and a user's choice of DIAMOND or BLASTX for its
``fine search'' phase; MICA demonstrates nearly order-of-magnitude run-time gains over the highly-optimized DIAMOND, comparable to that of caBLASTP over BLASTP.

Compressive genomics \cite{loh2012compressive} has been
generalized and adapted to non-sequence spaces as well, and coined
``compressive omics''~\cite{yu2015entropy}.
One such example is chemogenomics.
Applying a compressive acceleration approach, Ammolite~\cite{yu2015entropy}
accelerates SMSD search by an average of 150x on the PubChem database.
Another example is 
esFragBag~\cite{yu2015entropy}, which clusters proteins based on the cosine distance or
Euclidean distance of their bag-of-words vectors, further accelerating FragBag's
running time by an average of 10x.

@@

\section{Acknowledgments}
